Empowering Domain Recursion in Symmetric Weighted First-Order Model Counting

### 2022-01-21 notes/notes.tex
First draft of describing Ref

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathrsfs}

@@

\begin{definition}[Notation]
  For any clause $c = (P, N, C, \delta_c)$, bijection $\beta\colon \Vars(c) \to V$ (for some set of variables $V$) and function $\gamma\colon \Doms(c) \to \mathscr{D}$, let $c \ast (\beta, \gamma) = d$ be the clause with all occurrences of any variable $v \in \Vars(c)$ in $P$, $N$, and $C$ replaced with $\beta(v)$ (so $\Vars(d) = V$) and $\delta_d\colon V \to \mathscr{D}$ defined as $\delta_d \coloneqq \gamma \circ \delta_c \circ \beta^{-1}$. In other words, $\delta_d$ is the unique function that makes the following diagram commute:
  \[
  \begin{tikzcd}
    \Vars(c) \ar[r, tail, two heads, "\beta"] \arrow{d}{\delta_c} & V = \Vars(d) \ar[d, dashed, "\exists!\delta_d"] \\
    \Doms(c) \arrow{r}{\gamma} & \mathscr{D}.
  \end{tikzcd}
  \]
\end{definition}

\begin{algorithm}[t]
  \caption{A recursive function for checking whether one can reuse the circuit for computing $\WMC(\phi)$ to compute $\WMC(\psi)$. Both $\phi$ and $\psi$ are formulas, and $\rho\colon \Doms(\phi) \protect\pfun \Doms(\psi)$ is a partial map.}
  \label{alg:recursion}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKw{KwRet}{yield}
  \SetKwFunction{identifyRecursion}{identifyRecursion}
  \SetKwFunction{findHistory}{traceAncestors}
  \SetKwFunction{generateMaps}{generateMaps}
  \SetKwFunction{constructDomainMap}{constructDomainMap}
  \SetKwData{suitable}{suitableBijection}
  \SetKwData{found}{foundConstraintRemoval}
  \Fn{\identifyRecursion{$\phi$, $\psi$, $\rho = \emptyset$, $\found = {\normalfont \texttt{false}}$}}{
    \lIf{$|\phi| \ne |\psi|$ {\bf or} $\#\phi \ne \#\psi$}{\Return{\normalfont \texttt{null}}}
    \If{$\phi = \emptyset$}{
      \lIf{$\found$}{\Return{$\rho$}}
      \Return{\normalfont \texttt{null}}\;
    }
    \ForEach{clause $c \in \phi$\label{line:for1}}{
      \ForEach{clause $d \in \psi$ such that $\#d = \#c$\label{line:for2}} {
        \ForEach{$(\beta, \gamma) \in \generateMaps{$c$, $d$, $\rho$}$ such that $c \ast (\beta, \gamma) = d$\label{line:generateMaps}}{
          $\found' \gets \found$\;
          $\suitable \gets \texttt{true}$\;
          \ForEach{$e \in \Doms(c)$\label{line:vars}}{
            $\found'' \gets \findHistory{$e$, $\gamma(e)$}$\;
            \If{$\found'' = {\normalfont \texttt{null}}$}{
              $\suitable \gets \texttt{false}$\;
              break\;
            }
            \lIf{$\found''$}{$\found' \gets \texttt{true}$}
          }
          \If{\suitable}{
            $\rho'' \gets \identifyRecursion{$\phi \setminus \{\, c \,\}$, $\psi \setminus \{\, d \,\}$, $\rho \cup \gamma$, $\found'$}$\;
            \lIf{$\rho'' \ne {\normalfont \texttt{null}}$}{\Return{$\rho''$}}
          }
        }
      }
      \Return{\normalfont \texttt{null}}\;
    }
  }
  \Fn{\generateMaps{$c$, $d$, $\rho$}}{
    \ForEach{bijection $\beta\colon \Vars(c) \to \Vars(d)$\label{line:bijection}}{
      $\gamma \gets \constructDomainMap{$\Vars(c)$, $\delta_c$, $\delta_d$, $\beta$, $\rho$}$\;
      \lIf{$\gamma \ne {\normalfont \texttt{null}}$}{\KwRet{$(\beta, \gamma)$}}
    }
  }
  \Fn{\constructDomainMap{$V$, $\delta_c$, $\delta_d$, $\beta$, $\rho$}}{
    $\gamma \gets \emptyset$\;
    \ForEach{$v \in V$}{
      \lIf{$\delta_c(v) \in \dom(\rho)$ {\bf and} $\rho(\delta_c(v)) \ne \delta_d(\beta(v))$}{\Return{\normalfont \texttt{null}}}
      \lIf{$\delta_c(v) \not\in \dom(\gamma)$}{$\gamma \gets \gamma \cup \{\, \delta_c(v) \mapsto \delta_d(\beta(v)) \,\}$}
      \lElseIf{$\gamma(\delta_c(v)) \ne \delta_d(\beta(v))$}{\Return{\normalfont \texttt{null}}}
    }
    \Return{$\gamma$}\;
  }
  \Fn{\findHistory{$d$, $e$}}{
    $\found \gets \texttt{false}$\;
    \While{$e \ne d$ {\bf and} $e \in {\normalfont \dom}(\pi)$}{
      \lIf{$e \in \mathscr{C}$}{$\found \gets \texttt{true}$}
      $e \gets \pi(e)$\;
    }
    \lIf{$e = d$}{\Return{\found}}
    \Return{\normalfont \texttt{null}}\;
  }
\end{algorithm}

@@

@@

  The relevant domains and the definition of $\pi$ is in \cref{fig:example}. Since $\#\phi = \#\psi$, and the formulas are non-empty, the algorithm proceeds with the for-loops on \cref{line:for1,line:for2,line:generateMaps}. Suppose $c$ in the algorithm refers to \cref{phi1}, and $d$ to \cref{psi1}. Since both clauses have three variables, function \texttt{generateMaps} has $3!=6$ bijections to check. Suppose the identity bijection is picked first. Then \texttt{constructDomainMap} is called with the following parameters:
  \begin{itemize}
  \item $V = \{\, X, Y, Z \,\}$,
  \item $\delta_c = \{\, X \mapsto a, Y \mapsto b, Z \mapsto b \,\}$,
  \item $\delta_d = \{\, X \mapsto a', Y \mapsto b^\bot, Z \mapsto b^\bot \,\}$,
  \item $\beta = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$.
  \end{itemize}
  Since $\delta_i(Y) = \delta_i(Z)$ for $i \in \{\, c, d \,\}$, \texttt{constructDomainMap} returns $\gamma = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Thus, \texttt{generateMaps} yields its first pair of maps $(\beta, \gamma)$ to \cref{line:generateMaps}. Furthermore, the pair satisfies the $c \ast (\beta, \gamma) = d$ condition.

  Suppose \cref{line:vars} selects $e = a$ first. Since $\pi(a') = a$, and $a' \in \mathscr{C}$, \texttt{traceAncestors($a$, $a'$)} returns \texttt{true}, and the recursive call to \texttt{identifyRecursion($\phi'$, $\psi'$, $\gamma$, true)} is made, where $\phi'$ and $\psi'$ are new formulas with one clause each: \cref{phi2} and \cref{psi2}, respectively.

  Again we have two non-empty formulas with equal hash codes, so \texttt{generateMaps} is called with $c$ set to \cref{phi2}, $d$ set to \cref{psi2}, and $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Suppose \cref{line:bijection} picks the identity bijection first again.

\end{example}

TODO: conclude with a description of the inference rule and the node/edge type.

\section{New Node Types}

>>>>

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathrsfs}
\usepackage{amsfonts}

@@

\begin{definition}[Notation]
  For any clause $c = (P, N, C, \delta_c)$, bijection $\beta\colon \Vars(c) \to V$ (for some set of variables $V$) and function $\gamma\colon \Doms(c) \to \mathscr{D}$, let $c \ast (\beta, \gamma) = d$ be the clause with all occurrences of any variable $v \in \Vars(c)$ in $P$, $N$, and $C$ replaced with $\beta(v)$ (so $\Vars(d) = V$) and $\delta_d\colon V \to \mathscr{D}$ defined as $\delta_d \coloneqq \gamma \circ \delta_c \circ \beta^{-1}$. In other words, $\delta_d$ is the unique function that makes the following diagram commute:
  \[
  \begin{tikzcd}
    \Vars(c) \ar[r, tail, two heads, "\beta"] \arrow[d, swap, "\delta_c"] & V = \Vars(d) \ar[d, dashed, "\exists!\delta_d"] \\
    \Doms(c) \ar[r, swap, "\gamma"] & \mathscr{D}.
  \end{tikzcd}
  \]
\end{definition}

\begin{algorithm}[t]
  \caption{A recursive function for checking whether one can reuse the circuit for computing $\WMC(\phi)$ to compute $\WMC(\psi)$. Both $\phi$ and $\psi$ are formulas, and $\rho\colon \Doms(\phi) \protect\pfun \Doms(\psi)$ is a partial map.}
  \label{alg:recursion}
  \SetKwProg{Fn}{Function}{:}{}
  \SetKw{KwRet}{yield}
  \SetKwFunction{identifyRecursion}{identifyRecursion}
  \SetKwFunction{findHistory}{traceAncestors}
  \SetKwFunction{generateMaps}{generateMaps}
  \SetKwFunction{constructDomainMap}{constructDomainMap}
  \SetKwData{suitable}{suitableBijection}
  \SetKwData{found}{foundConstraintRemoval}
  \Fn{\identifyRecursion{$\phi$, $\psi$, $\rho = \emptyset$, $\found = {\normalfont \texttt{false}}$}}{
    \lIf{$|\phi| \ne |\psi|$ {\bf or} $\#\phi \ne \#\psi$}{\Return{\normalfont \texttt{null}}}
    \If{$\phi = \emptyset$}{
      \lIf{$\found$}{\Return{$\rho$}}
      \Return{\normalfont \texttt{null}}\;
    }
    \ForEach{clause $c \in \phi$\label{line:for1}}{
      \ForEach{clause $d \in \psi$ such that $\#d = \#c$\label{line:for2}} {
        \ForEach{$(\beta, \gamma) \in \generateMaps{$c$, $d$, $\rho$}$ such that $c \ast (\beta, \gamma) = d$\label{line:generateMaps}}{
          $\found' \gets \found$\;
          $\suitable \gets \texttt{true}$\;
          \ForEach{$e \in \Doms(c)$\label{line:vars}}{
            $\found'' \gets \findHistory{$e$, $\gamma(e)$}$\;
            \If{$\found'' = {\normalfont \texttt{null}}$}{
              $\suitable \gets \texttt{false}$\;
              break\;
            }
            \lIf{$\found''$}{$\found' \gets \texttt{true}$}
          }
          \If{\suitable}{
            $\rho'' \gets \identifyRecursion{$\phi \setminus \{\, c \,\}$, $\psi \setminus \{\, d \,\}$, $\rho \cup \gamma$, $\found'$}$\;\label{line:recursion}
            \lIf{$\rho'' \ne {\normalfont \texttt{null}}$}{\Return{$\rho''$}}
          }
        }
      }
      \Return{\normalfont \texttt{null}}\;
    }
  }
  \Fn{\generateMaps{$c$, $d$, $\rho$}}{
    \ForEach{bijection $\beta\colon \Vars(c) \to \Vars(d)$\label{line:bijection}}{
      $\gamma \gets \constructDomainMap{$\Vars(c)$, $\delta_c$, $\delta_d$, $\beta$, $\rho$}$\;
      \lIf{$\gamma \ne {\normalfont \texttt{null}}$}{\KwRet{$(\beta, \gamma)$}}
    }
  }
  \Fn{\constructDomainMap{$V$, $\delta_c$, $\delta_d$, $\beta$, $\rho$}}{
    $\gamma \gets \emptyset$\;
    \ForEach{$v \in V$}{
      \lIf{$\delta_c(v) \in \dom(\rho)$ {\bf and} $\rho(\delta_c(v)) \ne \delta_d(\beta(v))$}{\Return{\normalfont \texttt{null}}}
      \lIf{$\delta_c(v) \not\in \dom(\gamma)$}{$\gamma \gets \gamma \cup \{\, \delta_c(v) \mapsto \delta_d(\beta(v)) \,\}$}
      \lElseIf{$\gamma(\delta_c(v)) \ne \delta_d(\beta(v))$}{\Return{\normalfont \texttt{null}}}
    }
    \Return{$\gamma$}\;
  }
  \Fn{\findHistory{$d$, $e$}}{
    $\found \gets \texttt{false}$\;
    \While{$e \ne d$ {\bf and} $e \in {\normalfont \dom}(\pi)$}{
      \lIf{$e \in \mathscr{C}$}{$\found \gets \texttt{true}$}
      $e \gets \pi(e)$\;
    }
    \lIf{$e = d$}{\Return{\found}}
    \Return{\normalfont \texttt{null}}\;
  }
\end{algorithm}

@@

Diagramatically, \texttt{constructDomainMap} attempts to find $\gamma\colon \Doms(c) \to \Doms(d)$ such that the following diagram commutes (and returns \texttt{null} if such a function does not exist):
\[
\begin{tikzcd}
  \Vars(c) \ar[r, tail, two heads, "\beta"] \arrow[d, swap, "\delta_c"] & \Vars(d) \ar[d, "\delta_d"] \\
  \Doms(c) \ar[r, dashed, "\gamma"] \ar[d, hookrightarrow] & \Doms(d) \ar[d, hookrightarrow] \\
  \mathscr{D} \ar[r, swap, "\rho"] & \mathscr{D}.
\end{tikzcd}
\]

@@

  The relevant domains and the definition of $\pi$ is in \cref{fig:example}. Since $\#\phi = \#\psi$, and the formulas are non-empty, the algorithm proceeds with the for-loops on \cref{line:for1,line:for2,line:generateMaps}. Suppose $c$ in the algorithm refers to \cref{phi1}, and $d$ to \cref{psi1}. Since both clauses have three variables, in the worst case, function \texttt{generateMaps} would have $3!=6$ bijections to check. Suppose the identity bijection is picked first. Then \texttt{constructDomainMap} is called with the following parameters:
  \begin{itemize}
  \item $V = \{\, X, Y, Z \,\}$,
  \item $\delta_c = \{\, X \mapsto a, Y \mapsto b, Z \mapsto b \,\}$,
  \item $\delta_d = \{\, X \mapsto a', Y \mapsto b^\bot, Z \mapsto b^\bot \,\}$,
  \item $\beta = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
  \item $\rho = \emptyset$.
  \end{itemize}
  Since $\delta_i(Y) = \delta_i(Z)$ for $i \in \{\, c, d \,\}$, \texttt{constructDomainMap} returns $\gamma = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Thus, \texttt{generateMaps} yields its first pair of maps $(\beta, \gamma)$ to \cref{line:generateMaps}. Furthermore, the pair satisfies the $c \ast (\beta, \gamma) = d$ condition.

  Since $\pi(a') = a$, and $a' \in \mathscr{C}$, \texttt{traceAncestors($a$, $a'$)} returns \texttt{true}, which sets $\textsf{foundConstraintRemoval}'$ to \texttt{true} as well. When $e = b$, however, \texttt{traceAncestors}($b$, $b^\bot$) returns \texttt{false} since $b^\bot$ is a descendant of $b$ but not created by the constraint removal compilation rule. On \cref{line:recursion}, a recursive call to \texttt{identifyRecursion($\phi'$, $\psi'$, $\gamma$, true)} is made, where $\phi'$ and $\psi'$ are new formulas with one clause each: \cref{phi2} and \cref{psi2}, respectively.

  Again we have two non-empty formulas with equal hash codes, so \texttt{generateMaps} is called with $c$ set to \cref{phi2}, $d$ set to \cref{psi2}, and $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$. Suppose \cref{line:bijection} picks the identity bijection first again. Then \texttt{constructDomainMap} is called with the following parameters:
  \begin{itemize}
  \item $V = \{\, X, Y, Z \,\}$,
  \item $\delta_c = \{\, X \mapsto a, Y \mapsto b, Z \mapsto a \,\}$,
  \item $\delta_d = \{\, X \mapsto a', Y \mapsto b^\bot, Z \mapsto a' \,\}$,
  \item $\beta = \{\, X \mapsto X, Y \mapsto Y, Z \mapsto Z \,\}$,
  \item $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$.
  \end{itemize}
  Since $\beta$ and $\rho$ `commute' (TODO: as in the diagram above), and there are no new domains in $\Doms(c)$ and $\Doms(d)$, $\gamma$ exists and is equal to $\rho$. Again, the returned pair $(\beta, \gamma)$ satisfies the condition $c \ast (\beta, \gamma) = d$. This $\gamma$ passes the \texttt{traceAncestors} checks exactly the same way as the one before, and \cref{line:recursion} calls \texttt{identifyRecursion($\emptyset$, $\emptyset$, $\rho$, true)}, which immediately returns $\rho = \{\, a \mapsto a', b \mapsto b^\bot \,\}$ as the final answer. This means that one can indeed use a circuit for $\WMC(\phi)$ to compute $\WMC(\psi)$ by replacing every mention of $a$ with $a'$ and every mention of $b$ with $b^\bot$.
\end{example}

\subsection{Evaluation}

This $\rho$ is saved as a parameter of the new circuit node. During inference, there is a domain size map $\sigma\colon \mathscr{D} \to \mathbb{N}_0$, which gets updated to
\[
\sigma'(x) =
\begin{cases}
  \sigma(\rho(x)) & \text{if } x \in \dom(\rho) \\
  \sigma(x) & \text{otherwise}
\end{cases}
\]
for all $x \in \mathscr{D}$.

\section{New Operations}

TODO: compilation rules, node types, their evaluation during inference.
